Segmentation-registration of diffusion MR images of the brain driven by active-contours

### 2015-02-17 Publications/2015-NeuroImage/03-introduction.tex
finishing details of introduction

Nonlinear registration of images is a challenging task, ubiquitous
  in an endless number of image analysis applications.
The process aims to find a spatial mapping $U$ that aligns the information available
  in two different coordinate systems (generally called
  \emph{reference}, $R$, and \emph{moving}, $M$):
  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classify the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are those classified as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, measures distances between features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} combine properties of the iconic and geometric methods.
As reviewed in \citep{sotiras_deformable_2013}, an important research effort has been devoted to
  study application-oriented matching criteria.

@@

A fundamental antecedent of the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align \gls*{t1} and \gls*{dmri} images
  of the brain using a linear transformation model.
Similarly, we exploit the precise structures that can be extracted with
  widely-used software (i.e. \emph{FreeSurfer}~\citep{fischl_freesurfer_2012}),
  and then map these surfaces into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \cite{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving coordinate system.

>>>>

Nonlinear registration of images is a challenging task, ubiquitous
  in an endless number of image analysis applications.
The process aims to find a spatial mapping $U$ that aligns the information available
  in two different coordinate systems (generally called
  \emph{reference}, $R$, and \emph{moving}, $M$):
  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classify the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are those classified as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, align geometrical features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} map geometrical features and voxel-wise information.

@@

A fundamental antecedent of the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align \gls*{t1} and \gls*{dmri} images
  of the brain using a linear transformation model.
Similarly, we exploit the precise structures that can be extracted with
  widely-used software (i.e. \emph{FreeSurfer}~\citep{fischl_freesurfer_2012}),
  and then map these surfaces into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving coordinate system.

### 2015-02-17 Publications/2015-NeuroImage/04-methods.tex
finishing details of introduction

\section*{Methods}
\label{sec:methods}
%
\subsection*{Related work}
\label{sec:methods_background}
Registration-segmentation methods are generally built including a deformation model to
  support the evolution of a segmentation algorithm that drives the mapping process.
Thus, solutions usually select first a segmentation strategy to derive the cost function and
  regularization of the problem.
Generally, evolving active contours is the most extended segmentation approach to be
  integrated within the registration process.
These methods either have an implicit description of the relative boundary locations of the
  objects to be delineated using level sets \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005}, or model the statistical deviations from original shape
  \citep{cremers_kernel_2006,gastaud_combining_2004}.
Some Bayesian approaches have been proposed as well \citep{wyatt_map_2003,gass_simultaneous_2014}.
In \autoref{sec:methods_map} we start from a probabilistic framework to finally propose
	a cost function defined on the location of explicit surfaces, showing the duality of the
	existing approaches.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Soonly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping
  (the so-called ``2 \glspl*{pde} approach'').
\cite{vemuri_joint_2003} proposed the first atlas-based registration based on the level sets
  framework and using only one \gls*{pde}.
The principal application of these techniques was the image registration between timesteps in
  a time-series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} extended the idea of \citep{yezzi_variational_2001} to nonlinear
  registration using a free deformation field model.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on the level sets approximation.

\todo[inline]{Revise these references and include if necessary:
  \citep{pohl_unifying_2005,pohl_bayesian_2006,wang_joint_2006,guyader_combined_2011}}

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 15 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend \gls*{t1}, \gls*{t2} and multi-shell \gls*{dmri} images
  already corrected for artifacts, brain-extracted and spatially normalized,
  along with the most prominent results of the standard processing with
  \emph{FreeSurfer}: \emph{aparc} segmentations, and the surfaces corresponding to
  the cortical \gls*{gm} exterior and the \gls*{wm} interface with
  cortical \gls*{gm}.

>>>>

\section*{Methods}
\label{sec:methods}
%
\subsection*{Related work}
\label{sec:methods_background}
Registration-segmentation methods are generally built including a deformation model to
  support the evolution of a segmentation algorithm that drives the mapping process.
Thus, solutions usually select first a segmentation strategy to derive the cost function and
  regularization of the problem.
Generally, evolving active contours is the most extended segmentation approach to be
  integrated within the registration process.
These methods either have an implicit description of the relative boundary locations of the
  objects to be delineated using level sets \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005}, or model the statistical deviations from original shape
  \citep{cremers_kernel_2006,gastaud_combining_2004}.
Some Bayesian approaches have been proposed as well
  \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.
In \autoref{sec:methods_map} we start from a probabilistic framework to finally propose
	a cost function defined on the location of explicit surfaces, showing the duality of the
	existing approaches.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration using a level sets
  framework and only one \gls*{pde} for first.
The principal application of these techniques was the image registration between timesteps in
  a time-series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration with a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on the level sets approximation in the application of
  atlas-based segmentation.
A recent solution of the segmentation-registration problem close to the one proposed here was
  presented by \cite{guyader_combined_2011}.
Our method differs on two key features:
  it allows multiple active contours for segmentation, and the regularization strategy.
Whereas \cite{guyader_combined_2011} proposed a nonlinear elasticity smoother, we rely on
  the smoothing effects of the deformation field and the anisotropic regularization
  proposed by \cite{nagel_investigation_1986}.

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 15 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing with
  \emph{FreeSurfer}.

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures;
  3) cerebral \gls*{wm};
  4) brain stem and cerebellar \gls*{wm};
	5) cerebellar \gls*{gm}; and
	6) cortical \gls*{gm} surface.
The choice of this particular model of homogeneous regions responds to the joint distribution
  of \gls*{fa} and \gls*{adc} of the human brain, as shown in the {\color{red} Supplemental
  Materials (Figure XX)}.

In this case, we derived the deformation $U_{true}^{-1}$ from the field maps released with
  the corresponding packages of each dataset from the \gls*{hcp},
  mimicking the real distortions, using a derivation of our previous work
  \citep{esteban_simulationbased_2014}.
After warping the original \gls*{dmri} with $U_{true}^{-1}$, we computed the \gls*{dti} and
  the derived scalars (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.

### 2015-02-17 Publications/2015-NeuroImage/05-results.tex
finishing details of introduction

\subsection*{Distortion correction on real datasets}
\label{sec:results_hcp}

As we propose the \gls*{fa} and the \gls*{md} as target features, we
  implemented a simplified pipeline for processing \gls*{dmri} using
  \emph{MRtrix} \citep{tournier_mrtrix_2012}.
\todo[inline]{insert here the summary of parameters}

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract priors of the following
  homogeneous regions in terms of \gls*{fa} and \gls*{md} joint distribution:
\begin{itemize}
	\item \gls*{csf} of the ventricular system,
	\item deep \gls*{gm} structures,
	\item \gls*{wm} surface as the ensemble of brain stem, and
	  cerebellar and cerebral \gls*{wm},
	\item cortical \gls*{gm} surface, including cerebellar \gls*{gm}.
\end{itemize}

>>>>

\subsection*{Correction of distortions on real datasets}
\label{sec:results_hcp}

### 2015-02-17 Publications/2015-NeuroImage/06-discussion.tex
finishing details of introduction

% -*- root: 00-main.tex -*-
\section*{Discussion}
\label{sec:discussion}
We propose a registration-segmentation algorithm designed for biomedical
  image analyses that are affected by all or some of the following circumstances:
  \begin{itemize}
  	\item Reference and target images do not present appropriate contrasts to
  	perform volume-based registration reliably, or there is not available an apt image
 		to be used as reference.
  	\item The low resolution of the target volume produces strong partial volume effects
  	that exclude the use of voxel-wise segmentation algorithms and hinder volume-based
  	registration.
  	\item The unknown warping is smooth enough to be represented with a sufficient number
  	of BSpline kernels, and its explicit need is justified (i.e. the distortion
  	field in \gls*{dmri} images, object tracking applications, etc.).
  \end{itemize}
The algorithm is designed for the proposed application on correcting distortions of
  \gls*{dmri} data, a required preliminary step to extract the structural connectivity
  networks.
As a consequence, our approach relies on the availability of precise segmentations or
  surfaces of the objects that are to be segmented on the target volume(s).
Connectome extraction protocols generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
\todo[inline]{Add sentence about comment RW\#1 of IPMI2013: incorporate knowledge of
the EPI distortions into regularization}
Our method solves in a single step a joint problem usually addressed in two steps.
The first solves the linear registration of the \gls*{t1} and \gls*{dmri} data (i.e.
  \cite{greve_accurate_2009}), whereas the second corrects for the nonlinear distortions
  derived from the inhomogenety of magnetic susceptibility across tissues
  \citep{jezzard_correction_1995}.

{\color{red} We also considered the reproducibility of results a design requirement.
Therefore, real data are fetched from a publicly available repository
  (the Human Connectome Project \citep{essen_human_2012}) and all the software
  involved in this paper is also publicly released.
\autoref{fig:evworkflows} describes the general structure of the workflow we implemented
  as evaluation instrument, using \emph{nipype} \citep{gorgolewski_nipype_2011} to ensure
  reproducibility.}

We implemented the registration method upon the widely used Insight Registration and Segmentation
	Toolkit (\url{http://www.itk.org}) with the aim to release a useful and free software bundle.
The instrumentation framework is also distributed with the registration method,
  and has been implemented using \emph{nipype} \citep{gorgolewski_nipype_2011}.
The pipelines include the automatic generation of the ``cortex'' phantom.
The real datasets are publicly available under the \gls*{hcp} project.
Thus, all the experiments and results presented in this paper are intended to be
  fully reproducible.

>>>>

% -*- root: 00-main.tex -*-
\section*{Discussion}
\label{sec:discussion}
We propose a joint registration and segmentation algorithm designed for biomedical
  image analyses that are affected by all or some of the following circumstances:
  i) reference and target images do not present appropriate contrasts to
  	perform volume-based registration reliably, or there is not available an apt image
 		to be used as reference;
  ii) the low resolution of the target volume produces strong partial volume effects
  	that exclude the use of voxel-wise segmentation algorithms and hinder volume-based
  	registration; and,
  iii) the estimation of the unknown warping is explicitly needed and the
    deformation field is smooth enough to be represented with a sufficient number
  	of B-Spline kernels (i.e. the distortion in \gls*{dmri} images,
  	object tracking in time series, etc.).
The registration method performs the segmentation of the target image in homogeneous
  and competing regions defined by explicit active contours.
We chose active contours without edges as we were interested on segmenting homogeneous
  regions of diffusion properties within the brain.
Finally, a B-Spline deformation field supports and keeps track of the evolution of the
  active contours.
Iteratively, the active contours are updated following the instant shape gradients
  \citep{herbulot_segmentation_2006} with a gradient descent strategy.
In section \nameref{sec:methods_background}, we framed our method within the current
  state-of-art and defined the novelty of our approach.

To support our claim that images can be simultaneously segmented and registered to a
  reference coordinate system, we conducted evaluation experiments on digital
  simulated phantoms and real datasets.

The algorithm is designed for the proposed application on correcting distortions of
  \gls*{dmri} data, a required preliminary step to extract the structural connectivity
  networks.
As a consequence, our approach relies on the availability of precise segmentations or
  surfaces of the objects that are to be segmented on the target volume(s).
Connectome extraction protocols generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
\todo[inline]{Add sentence about comment RW\#1 of IPMI2013: incorporate knowledge of
the EPI distortions into regularization}
Our method solves in a single step a joint problem usually addressed in two steps.
The first solves the linear registration of the \gls*{t1} and \gls*{dmri} data (i.e.
  \cite{greve_accurate_2009}), whereas the second corrects for the nonlinear distortions
  derived from the inhomogenety of magnetic susceptibility across tissues
  \citep{jezzard_correction_1995}.

\paragraph*{Availability and reproducibility statement}
{\color{red} We also considered the reproducibility of results a design requirement.
Therefore, real data are fetched from a publicly available repository
  (the Human Connectome Project \citep{essen_human_2012}) and all the software
  involved in this paper is also publicly released.
\autoref{fig:evworkflows} describes the general structure of the workflow we implemented
  as evaluation instrument, using \emph{nipype} \citep{gorgolewski_nipype_2011} to ensure
  reproducibility.

We implemented the registration method upon the widely used Insight Registration and Segmentation
	Toolkit (\url{http://www.itk.org}) with the aim to release a useful and free software bundle.
The instrumentation framework is also distributed with the registration method,
  and has been implemented using \emph{nipype} \citep{gorgolewski_nipype_2011}.
The pipelines include the automatic generation of the ``cortex'' phantom.
The real datasets are publicly available under the \gls*{hcp} project.
Thus, all the experiments and results presented in this paper are intended to be
  fully reproducible.}

### 2015-02-17 Publications/2015-NeuroImage/08-acknowledgements.tex
finishing details of introduction

% -*- root: 00-main.tex -*-
% use section* for acknowledgement
\section*{Acknowledgment}
The authors thank V. Estellers for early discussions at the beginning of this project,
  and L. Vese for her support during OE's research visits in her laboratory.

DZ is supported by the Swiss National Science Foundation (SNF) under grant PBELP2-137727.

This study is supported by: the Spain's Ministry of Science and Innovation
(projects TEC2010-21619- C04-03, TEC2011-28972-C02-02, CDTI-CENIT
AMIT and INNPACTO PRECISION), Comunidad de Madrid (ARTEMIS) and
European Regional Development Funds; the Center for Biomedical Imaging
(CIBM) of the Geneva and Lausanne Universities and the EPFL, as well as the
Leenaards and Louis Jeantet foundations.

>>>>

% -*- root: 00-main.tex -*-
% use section* for acknowledgement
\section*{Author contributions}
All the authors contributed in editing the paper.
OE implemented the method, designed and conducted the experiments, wrote the paper,
  simulated the phantoms and prepared the real data.
DZ devised and drafted the registration method, generated early phantom datasets, and
  collaborated with the implementation of the method.
AD, MBC and MJLC interpreted the results.
AD, MBC, MJLC, JPT and AS advised on all aspects of the work.

\section*{Acknowledgment}
The authors thank V. Estellers for early discussions at the beginning of this project,
  and L. Vese for her support during OE's research visits in her laboratory.

This study is supported by: the Spain's Ministry of Science and Innovation
  (projects TEC2010-21619- C04-03, TEC2011-28972-C02-02, CDTI-CENIT
  AMIT and INNPACTO PRECISION), Comunidad de Madrid (ARTEMIS) and
  European Regional Development Funds; the Center for Biomedical Imaging
  (CIBM) of the Geneva and Lausanne Universities and the EPFL, as well as the
  Leenaards and Louis Jeantet foundations.
DZ is supported by the Swiss National Science Foundation (SNF) under grant PBELP2-137727.
